---
title: "Homework 1"
author: "Vladislav Zakatov"
date: "15 October 2015"
output: pdf_document
---

This document has been created as part of the first homework assignment at CMF.

Initially, the working directory is set and the required *quantmod* package is loaded.

```{r results='hide', message=FALSE}
##### Initialisation #####
setwd("~/")
Sys.setlocale("LC_ALL","English")
library(quantmod)
```

The historical prices of S&P 500 index, Apple and IBM from 2002 up to present are loaded from Yahoo Finance using *getSymbols* function from *quantmod* package. The raw data frames are then merged into a single xts using only adjusted prices.

```{r results='hide'}
##### Loading data #####
tickers = c("SPY", "AAPL", "IBM") # specify the tickers to load

e <- new.env() # environment where data will be stored

# suppress warning about future change in the function behaviour
options("getSymbols.warning4.0"=FALSE) 

# load data from Yahoo Finance
getSymbols(tickers, src = "yahoo", from="2002-01-01", env=e) 

# merge data into a single xts taking only adjusted prices
df = do.call(merge, eapply(e, Ad)[tickers]) 
names(df) = tickers
```

Daily returns are calculated and saved as *xts*.

```{r}
##### Calculate returns #####
rets = apply(df, 2, function(x) diff(x)/x[-length(x)])
rets = xts(rets, order.by = index(df)[-1])
```

Both prices and returns are then plotted. 

```{r fig.width = 12, fig.height = 17}
##### Plot prices and returns #####
for (i in 1:ncol(df))
{
  par(mfrow=c(2,1))
  plot(df[,i], main = paste(tickers[i], "price"))
  plot(rets[,i], main = paste(tickers[i], "returns"))
  par(mfrow=c(1,1))
}
```

The user-defined function *stats* calculates mean and standard deviation of the returns series which is passed as an argument to the function.

```{r}
###### Calculate mean return and return s.d. #####
stats = function(r)
{
  x = c(mean(r), sd(r))
  names(x) = c("Mean", "SD")
  return(x)
}

options(digits=4)
apply(rets, 2, stats)
```